### 2016-08-28 Chapters/ContinuumSolvationModels.tex
Work on introduction
Reshuffle some sections and dump some reference around

@@

\section[Variational Formulation of Classical Polarizable Models]{
Classical Polarizable Models within a Variational Formulation}
\label{sec:variational}

>>>>

\section{Continuum Solvation Models: Why and How}\label{sec:CSM-why-how}
\todo[inline]{CURRENTLY VERBATIM FROM MASTER THESIS}

Continuum models arise as limits, in a proper sense, to the discrete
models. To stress this point we introduce the \emph{solution}
Hamiltonian in the Born--Oppenheimer approximation as:
\begin{equation}\label{eq:solution-ham}
 H(\vect{r}_\mathrm{M}, \vect{r}_\mathrm{S}) =
  H_\mathrm{M}(\vect{r}_\mathrm{M}) +  H_\mathrm{S}(\vect{r}_\mathrm{S})
+ H_\mathrm{MS}(\vect{r}_\mathrm{M}, \vect{r}_\mathrm{S})
\end{equation}
where the subscript M refers to the solute while S to the solvent. The
coordinates $(\vect{r}_\mathrm{M}, \vect{r}_\mathrm{S})$ refer to
\emph{both} nuclei and electrons, the interaction term is given by the
usual Coulombic electrostatic Hamiltonian.

It is important to remark that nuclear coordinates must be considered
explicitly. Liquid phases are, by definition, characterized by the high
mobility of their molecular constituents. Changes in the internal
geometry of solvent molecules, triggered by intermolecular interactions
between themselves and with the solute, always play a role in the
solvation process. Whether this role is prominent or not cannot be
decided \emph{a priori}.

From a computational point of view, using the Hamiltonian in Eq.~\eqref{eq:solution-ham} is simply not feasible. We would have to
determine a \acs{PES} of enormous dimensionality and then explore the full
dynamics of the system. Even assuming that the minimum energy
conformation on the \acs{PES} can be fully characterized, its knowledge will
not lead us to a correct description of the physics of solvation. In
fact, macroscopic observables of systems with a large number of degrees
of freedom arise as \emph{average} values of the microscopic behaviour
of every single component, the average being either on the phase space
trajectory or on the appropriate ensemble.~\autocite{Hill1960-ql}. It is always
necessary to perform some kind of average upon the energies
corresponding to all the accessible conformations of the whole
solute+solvent system.

Of course, this formidable task may be accomplished using statistical
thermodynamics and the idea of \emph{statistical ensembles} developed by
Gibbs. To apply the idea of an ensemble average we must resort to a
molecular dynamics or Monte Carlo simulation of the solution, in order
to obtain the appropriate partition function, $Z$. This introduces a
number of points that will be central in the development of continuum
models:
\begin{enumerate}
 \item statistical simulations rely on \emph{macroscopic} parameters
   such as the temperature or the density. These parameters are not
   present in the Hamiltonian \eqref{eq:solution-ham}, but are necessary
   for the description of molecular systems in the condensed phase. This
   is true whether one considers discrete or continuum approaches;
 \item the basic energetic quantity is the thermodynamic potential
   characterizing the chosen ensemble. If a direct comparison with the
   experiment is sought, one should run a simulation in the $(NpT)$
   ensemble whose potential is the Gibbs free energy $G$. This point is
   of paramount importance in continuum models and we will have the
   opportunity to stress it in this Chapter as well as in Chapter
   \ref{chap:relpcmscf};
 \item even starting from a completely atomistic point of view, our
   final results will always be in the form of an average over solvent
   conformations. We have explicitly replaced a discrete description
   with a continuous distribution function. Atomistic details can of
   course be recovered by further analysis, but the point remains:
   properties of solutions (the solvation energy among them) have an
   intrinsic average nature.
\end{enumerate}

The fact that an averaging procedure is essential in studying solvation
processes should convince the reader that, at least from a purely
mathematical point of view, we can replace the full Hamiltonian
\eqref{eq:solution-ham} with an \emph{effective} Hamiltonian, averaging
the solvent degrees of freedom in a preliminary
step.~\autocite{Angyan1992-vo, Tapia1992-pu}
\todo[inline]{Some pieces missing here, FILL IN}
Alternatively, an approximate decoupling can be obtained replacing the
mean-field operator in Eq.~\eqref{eq:ave1} by its classical average over
a suitable ensemble:
\begin{equation}\label{eq:ave_classic}
 \Braket{H_\mathrm{MS}}_\mathrm{S} = \frac{\int\diff\Gamma H_\mathrm{MS}\exp
\left(-\beta H_\mathrm{S}\right)}{\int\diff\Gamma\exp\left(-\beta H_\mathrm{S}\right)}.
\end{equation}
This classical average can be written in terms of the $g(\vect{r})$ distribution functions. These functions are defined
for the electrons and the nuclei of the solvent with respect to the fixed frame of reference of the solute molecule, see \noparcite[ref.][]{hansen}.
This amounts to considering the solvent as a structureless continuum, hence defining continuum models.

Continuum models are then characterized, by a solute-only
\emph{effective} Hamiltonian:\footnote{We will use the symbol
$H_\mathrm{MS}$ for the interaction Hamiltonian even if it is not the
full interaction term of Eq.~\eqref{eq:solution-ham}.}
\begin{equation}\label{eq:effective-ham}
 H(\vect{r}_\mathrm{M}) = H_\mathrm{M}(\vect{r}_\mathrm{M})
+ H_\mathrm{MS}(\vect{r}_\mathrm{M}),
\end{equation}
the explicit form of the averaged interaction in
Eq.~\eqref{eq:effective-ham} is in terms of solvent \emph{response
functions}, $Q_x(\vect{r},\vect{r}^\prime)$.
As will be apparent in the following, this term may depend on the solute
density, making the effective Hamiltonian \emph{nonlinear}.

An hierarchy of approximations for the resolution of the effective
Schr\"odinger equation can be devised, as was done for the discrete
models. The distinguishing feature is now the \emph{order} of the
solvent response functions used in the definition of the interaction
term.

If we limit ourselves to a linear response of the solvent we obtain the
class of models commonly called continuum solvation models, the
Polarizable Continuum Model among them. Computer simulations can be used
to determine these functions, but almost always one uses bulk properties
of the pure solvent obtained from experiment.

This, at first sight, crude approximation is rather well justified by
the following argument: from a physical point of view, our solution is
extremely dilute. The presence of the solute is an extremely small
perturbation for the bulk of the solvent and it is safe to assume that
the macroscopic properties of the solution are not far from those of the
pure solvent. Furthermore, the use of macroscopic experimental values
implicitly takes into account the averaging procedure one would carry
out in atomistic simulations of the solution.

As pictorially represented in Figure \ref{fig:discretetocontinuum} the
solvent is now a \emph{structureless polarizable} continuum and the
solute is accommodated inside a cavity formed in this continuum.
\begin{figure}[!h]
 \centering
\includegraphics[width=.5\textwidth]{discretetocontinuum}
\caption{Pictorial representation of the transition from a discrete to a
continuum model of a solution.}
\label{fig:discretetocontinuum}
\end{figure}

@@

\section[Variational Formulation of Classical Polarizable Models]{
Above and Beyond: Classical Polarizable Models within a Variational Formulation}
\label{sec:variational}

### 2016-08-28 Chapters/QuantumMechanics.tex
Reshuffle some sections and dump some reference around

\begin{itemize}
  \item nonrelativistic and relativistic Hamiltonians used in molecular
    quantum mechanics
  \item \ac{BO} approximation and the concept of \ac{PES}
  \item Fock space
  \item single-reference methods: \ac{HF} and \ac{KS} \ac{DFT}
\end{itemize}

>>>>

\begin{itemize}
  \item nonrelativistic and relativistic Hamiltonians used in molecular
    quantum mechanics
  \item \ac{BO} approximation and the concept of \ac{PES}
  \item Fock space
  \item single-reference methods: \ac{HF} and \ac{KS} \ac{DFT}
  \item the \ac{CC}  model
\end{itemize}

\pagebreak

\section{A Brief Summary of Coupled Cluster Theory}

\todo[inline]{MOVED FROM CHAPTER 4 HERE: READJUST!}

The analyses and derivations in this Chapter are preeminently based
on the \acl{CC} wave function Ansatz.
We will assume a closed-shell, \acs{HF} single reference function and
construct our correlated wave function as an exponential mapping:
\begin{equation}
  \ket{\mathrm{CC}} = \expo{T}\ket{\mathrm{HF}}.
\end{equation}
The \emph{cluster} operator appearing in the exponential is given as:
\begin{equation}
  T = \sum_{p=1}^\mathcal{M} T_p = \sum_{p=1}^\mathcal{M}\sum_{\mu_p}\tamp{\mu_p}\cluster{\mu_p}
\end{equation}
that is, as a truncated sum of excitation operators, $\cluster{\mu_p}$,
times the corresponding cluster amplitudes, $\tamp{\mu_p}$.
Here $\mu_p$ is the $\mu$-th excitation at the $p$-th excitation level and
$\mathcal{M}$ is the truncation level.
One of the main strengths of the \acs{CC} model is its size extensivity
which stems directly from the exponential nature of the wave
operator.~\autocite{Shavitt2009-mr, Helgaker2000-tz}

The \acs{CC} method is a \emph{projective} method. Chosen the truncation
level $\mathcal{M}$, one projects the nonrelativistic Schrödinger
equation for this Ansatz on the excitation manifold which comprises the
reference function and all possible excited determinants, up to the
chosen truncation order:
\begin{subequations}
  \begin{align}
    \braket{\mathrm{HF} | H_0\expo{T} | \mathrm{HF}} &= E_\mathrm{CC} \\
    \braket{\mathrm{exc} | H_0\expo{T} | \mathrm{HF}} &=
    E_\mathrm{CC}\braket{\mathrm{exc} | \expo{T} | \mathrm{HF}}
  \end{align}
\end{subequations}
where $\bra{\mathrm{exc}}$ means any of the excited determinants in Fock
space compatible with the excitation level truncation.
This is the \emph{unlinked} form of the \acs{CC}
equations.~\autocite{Helgaker2000-tz} It is usually more convenient to
perform a \emph{similarity transformation} of the Hamiltonian operator,
leading to the \emph{linked} form of the \acs{CC} equations:
\begin{subequations}
  \begin{align}
    \braket{\mathrm{HF} | \expo{-T}H_0\expo{T} | \mathrm{HF}} &= E_\mathrm{CC} \\
    \braket{\mathrm{exc} | \expo{-T}H_0\expo{T} | \mathrm{HF}} &= 0
  \end{align}
\end{subequations}
The linked and unlinked formulations are completely equivalent. However,
the former is size extensive term-by-term.
Use of similarity transformed operators is ubiquitous when dealing with
\acs{CC} theory and we introduce the following notation for it:
\begin{equation}
  \overline{O} =\expo{-T}O\expo{T}
\end{equation}
The cluster operator is not Hermitian. Hence, the similarity transformation
will not, in general, preserve any of the symmetries, such as
hermiticity, that the bare operators might have possessed.
Similarity-transformed operators can be expanded in a commutator series,
also known as the \ac{BCH} expansion:
\begin{equation}
  \overline{O} =
  \sum_{n=0}^{\infty}\frac{1}{n!}[A, B]^n
\end{equation}
where $[A,B]^n$ is the $n$-fold nested commutator. It can be shown that
the \acs{BCH} expansion of the similarity-transformed Hamiltonian
\emph{exactly} truncates after the four-fold nested commutator, greatly
simplifying algebraic derivations and manipulations.

Introducing the \acl{MP} partitioning of the Hamiltonian leads to
further insight:
\begin{equation}
 H_0 = F + \Phi = \sum_P\epsilon_P\cons{P}\anni{P} + (g- V_\mathrm{HF})
\end{equation}
where $F$ if the Fock matrix and $\Phi$ is the fluctuation potential.
Given our initial assumption on the reference function, the Fock
operator is diagonal and expressed in terms of spin-orbital energies and
number operators.
Its similarity transformation truncates after the second term and has a
relatively compact form:
\begin{equation}
  \overline{F} = F + \sum_{i}^4\sum_{\mu_i}\tamp{\mu_i}\cluster{\mu_i}\denom{\mu_i}
\end{equation}
where $\denom{\mu_i}$ is the difference in orbital energies between the
virtual and occupied spin-orbitals of excitation $i$.
For the fluctuation potential the similarity transformation truncates
after the fifth term, including up to four-fold nested commutator.

As formulated so far, the \acs{CC} method is not variational. It is
however possible to introduce a fully variational Lagrangian that
leads to the same amplitudes equations. The Lagrangian is built by
treating the amplitude equations as constraints in the optimization of
the \acs{CC} energy:
\begin{equation}\label{eq:cc-lagrangian}
  \begin{aligned}
    \mathcal{L}^\mathcal{M}(\tamp{}, \tbar{})
    &=
    \braket{\mathrm{HF} | \overline{H_0} | \mathrm{HF}}
    + \sum_{i=1}^{\mathcal{M}}\sum_{\mu_i}
    \tbar{\mu_i}\braket{\mu_i | \overline{H_0} | \mathrm{HF}} \\
    &=
  E_0
  + \sum_{i=1}^{\mathcal{M}}\tbar{\mu_i}\denom{\mu_i}\tamp{\mu_i}
  + \braket{\mathrm{HF} | \overline{\Phi} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \overline{\Phi} | \mathrm{HF}}
  \end{aligned}
\end{equation}
where the shorthand notation for the Lagrangian multiplier state was
introduced:
\begin{equation}
  \bra{\tbar{j}} = \sum_{\mu_j}\tbar{\mu_j}\bra{\mu_j}
\end{equation}
and the \acs{MP}-partitioned form of the Hamiltonian was also exploited.
Differentiation of the Lagrangian with respect to the multipliers
correctly yields the amplitudes equations, while differentiation with
respect to the amplitudes leads to the governing equations for the
multipliers. While these are not needed for the evaluation of the
\acs{CC} energy, their calculation is mandatory when expectation values
and molecular properties in general are sought.
The \acs{CC} expectation values are formed using the left and
right \acs{CC} states and we introduce the following notation:
\begin{equation}
  O(\tamp{}, \tbar{}) = \braket{\mathrm{HF} | \overline{O} | \mathrm{HF} }
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \overline{O} | \mathrm{HF} }
\end{equation}
When required by the context, we will explicitly write at which
truncation level in the cluster operator the expectation was evaluated:
$O^\mathcal{M}(\tamp{}, \tbar{})$.

Having summarized the essential points of \acs{CC} theory and having
introduced the relevant notation, we can proceed to analyze its coupling
with quantum/classical polarizable Hamiltonians.

### 2016-08-28 Chapters/ResponseTheory.tex
Reshuffle some sections and dump some reference around

\begin{itemize}
    \item Something about exact state response theory: quasienergy,
      response functions, residues of response functions.~\autocite{}
    \item  Derivation of the linear response function with
      quantum/classical polarizable Hamiltonian and then refer to the
      OpenRSP paper for the open-ended treatment.~\autocite{pcm-openrsp}
\end{itemize}

>>>>

\begin{itemize}
    \item Something about exact state response theory: quasienergy,
      response functions, residues of response functions.
    \item  Derivation of the linear response function with
      quantum/classical polarizable Hamiltonian and then refer to the
      OpenRSP paper for the open-ended treatment.~\autocite{pcm-openrsp}
\end{itemize}

\pagebreak

\section{Exact State Response Theory}\label{sec:exact-response}

General theory~\autocite{Olsen1985-nr, Christiansen1998-pe,
Norman2011-ad, Helgaker2012-cz, Pawlowski2015-sq}
Solution strategy for response equations~\autocite{Kauczor2011-rd}
Open-ended \acs{AO}-based \acs{SCF} response~\autocite{
Larsen2000-hj, Kjaergaard2008-hy, Thorvaldsen2008-sg,
Kristensen2008-hv, Ringholm2014-gx}
Open-ended \acs{SCF} residues~\autocite{Friese2015-kb}

\section{SCF Response Theory for Quantum/Classical Polarizable Hamiltonians}\label{sec:csm-response}

\acs{PCM}~\autocite{Cammi1994-qj, Cammi1996-wf, Cammi1996-vx,
Cammi1999-rb, Cammi2003-qy, Frediani2005-nc, Ferrighi2010-pm}
\acs{FQ} and \acs{FQ}/\acs{PCM}~\autocite{Lipparini2012-hx, Lipparini2012-tl, Lipparini2013-ud}
\acs{PE}~\autocite{Olsen2010-wa}
\acs{QM}/\acs{MM}/Continuum~\autocite{Stenidal2011-ki, Caprasecca2012-ir}
Paper V~\autocite{pcm-openrsp}
